---
title: "PCA and MDS"
author: "Ramiro Barrantes-Reynolds"
date: "2026-09-09"
categories: [statistics, data visualization, science-communication]
subtitle: "A plain-language guide to two of the most common plots in a data report"
execute:
  echo: false
  message: false
---

## Introduction

Here we explain two multivariate methods: **PCA** (Principal Component Analysis) or **MDS** (Multidimensional Scaling). They are multivariate as we commonly have real data dozens, hundreds, or thousands of numbers attached to it — gene activity levels, epigenetic measurements, etc. 

What we do is to reduce all those numbers down to a simple map — usually just two or three dimensions while trying to preserve **which samples are similar to each other, and which are different.**

## PCA: Finding the Most Informative Angle

Imagine holding an oddly shaped object — a stretched, lumpy potato — and shining a flashlight on it to cast a shadow on the wall. From some angles, the shadow is a shapeless blob that tells you almost nothing about the object. From other angles, the shadow captures the object's real shape and spread remarkably well.

PCA is a systematic search for the best angle. It looks at all your original measurements and finds the "view" that preserves the most spread — the most real difference between samples — in as few dimensions as possible. That best angle becomes the first axis of the plot; the next-best (and unrelated) angle becomes the second axis. Those chosen angles are called **principal components**.

The plots below make this concrete with just two measurements. First, just the raw data: each dot is a sample with two correlated measurements — knowing one tells you a lot about the other, which is why the cloud runs diagonally instead of in a circle. Then, the two principal directions overlaid on that same cloud. Finally, PCA has simply *rotated* the cloud so its long direction lines up with the horizontal axis:

```{r}
#| label: fig-pca-rotation
#| fig-cap: "60 samples from two groups, before and after PCA. The red line marks the direction of most spread (PC 1); the green line, its perpendicular runner-up (PC 2). Left: the raw data alone, no lines. Middle: the same data, with those directions cut diagonally across it, right along the seam between the two groups. Right: the identical cloud, colors, and two lines, rotated so the red line becomes the x-axis and the green line becomes the y-axis."
#| fig-subcap:
#|   - "Original Measurements"
#|   - "Original Measurements, with PC Directions"
#|   - "Rotated to the Best Angle"
#| layout-ncol: 3

library(ggplot2)

set.seed(4)
n_per_group <- 30
offset <- 3.2

make_group <- function(offset) {
  x1 <- rnorm(n_per_group, 0, 1.6) + offset
  x2 <- 0.9 * (x1 - offset) + rnorm(n_per_group, 0, 0.6) + offset
  data.frame(x1 = x1, x2 = x2)
}

orig <- rbind(make_group(-offset), make_group(offset))
orig$group <- rep(c("Group 1", "Group 2"), each = n_per_group)

pca_2d <- prcomp(orig[, c("x1", "x2")], center = TRUE, scale. = FALSE)
rot <- data.frame(pc1 = pca_2d$x[, 1], pc2 = pca_2d$x[, 2], group = orig$group)

# the two principal directions, drawn as full lines through the original data
center <- pca_2d$center
rotation <- pca_2d$rotation
slope_pc1 <- rotation[2, 1] / rotation[1, 1]
slope_pc2 <- rotation[2, 2] / rotation[1, 2]
intercept_pc1 <- center[2] - slope_pc1 * center[1]
intercept_pc2 <- center[2] - slope_pc2 * center[1]

lim <- max(abs(c(orig$x1, orig$x2, rot$pc1, rot$pc2))) * 1.1
pal <- c("Group 1" = "#4e79a7", "Group 2" = "#e15759")

ggplot(orig, aes(x1, x2, color = group)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = pal) +
  coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(x = "Measurement 1", y = "Measurement 2", color = NULL) +
  theme_minimal()

ggplot(orig, aes(x1, x2, color = group)) +
  geom_abline(slope = slope_pc1, intercept = intercept_pc1, color = "firebrick", linewidth = 0.8) +
  geom_abline(slope = slope_pc2, intercept = intercept_pc2, color = "darkgreen", linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = pal) +
  coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(x = "Measurement 1", y = "Measurement 2", color = NULL) +
  theme_minimal()

ggplot(rot, aes(pc1, pc2, color = group)) +
  geom_hline(yintercept = 0, color = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = 0, color = "darkgreen", linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = pal) +
  coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
  labs(x = "PC 1", y = "PC 2", color = NULL) +
  theme_minimal()
```

Nothing about the samples changed between those two plots — they, and the two colored lines, were only rotated. The red line (PC 1) is the same "most spread" direction in both panels; it just happens to line up with the page's horizontal axis once PCA is done rotating.

Notice thought that, on the left we have that "Measurement 1" and "Measurement 2" meant something specific and nameable. On the right, PC 1 is some blend of both original measurements — a direction, not a thing you measured. The principal components are no longer interpretable in the same way.

Below is another example in 3-D.

```{r}
#| label: fig-pca-3d
#| fig-cap: "Two groups with three measurements each. Left: the raw 3D cloud, with its three principal directions drawn as dotted lines (red = PC 1, green = PC 2, purple = PC 3). Middle: the identical cloud and the identical three lines, rotated so they align exactly with the box's edges. Right: PC 3 dropped completely — just PC 1 vs. PC 2, flat on the page — and the groups are still cleanly separated."
#| fig-subcap:
#|   - "Original Measurements (3D)"
#|   - "Rotated to the Best Angle (3D)"
#|   - "PC 3 Dropped, Back to Flat 2D"
#| layout-ncol: 3

library(ggplot2)
library(lattice)

set.seed(4)
n_per_group <- 25
offset <- 3.5

gen3d <- function(offset) {
  x1 <- rnorm(n_per_group, 0, 1.3) + offset
  x2 <- 0.8 * (x1 - offset) + rnorm(n_per_group, 0, 0.5) + offset
  x3 <- 0.7 * (x1 - offset) + rnorm(n_per_group, 0, 0.5) + offset
  data.frame(x1 = x1, x2 = x2, x3 = x3)
}

orig3 <- rbind(gen3d(-offset), gen3d(offset))
group3 <- factor(rep(c("Group 1", "Group 2"), each = n_per_group))

pca3 <- prcomp(orig3, center = TRUE, scale. = FALSE)
rot3 <- as.data.frame(pca3$x)
names(rot3) <- c("pc1", "pc2", "pc3")
rot3$group <- group3

cam <- list(z = 30, x = -70, y = 0)

# a principal direction, traced as a dense line of small points through the cloud's center
axis_line <- function(center, direction, len = 7, n = 80) {
  t <- seq(-len, len, length.out = n)
  data.frame(
    x1 = center[1] + t * direction[1],
    x2 = center[2] + t * direction[2],
    x3 = center[3] + t * direction[3]
  )
}

grp_levels <- c("Group 1", "Group 2", "PC1 axis", "PC2 axis", "PC3 axis")
cols <- c("#4e79a7", "#e15759", "firebrick", "darkgreen", "purple")
pchs <- c(19, 19, 20, 20, 20)
cexs <- c(1.1, 1.1, 0.35, 0.35, 0.35)
pal <- c("Group 1" = "#4e79a7", "Group 2" = "#e15759")

# original space: the three principal directions cut diagonally through the data
lines_orig <- rbind(
  cbind(axis_line(pca3$center, pca3$rotation[, 1]), grp = "PC1 axis"),
  cbind(axis_line(pca3$center, pca3$rotation[, 2]), grp = "PC2 axis"),
  cbind(axis_line(pca3$center, pca3$rotation[, 3]), grp = "PC3 axis")
)
combined_orig <- rbind(cbind(orig3, grp = as.character(group3)), lines_orig)
combined_orig$grp <- factor(combined_orig$grp, levels = grp_levels)

# PC space: those same three directions are now just the coordinate axes
lines_rot <- rbind(
  cbind(axis_line(c(0, 0, 0), c(1, 0, 0)), grp = "PC1 axis"),
  cbind(axis_line(c(0, 0, 0), c(0, 1, 0)), grp = "PC2 axis"),
  cbind(axis_line(c(0, 0, 0), c(0, 0, 1)), grp = "PC3 axis")
)
names(lines_rot) <- c("pc1", "pc2", "pc3", "grp")
combined_rot <- rbind(cbind(rot3[, c("pc1","pc2","pc3")], grp = as.character(group3)), lines_rot)
combined_rot$grp <- factor(combined_rot$grp, levels = grp_levels)

cloud(x3 ~ x1 * x2, data = combined_orig, groups = grp, col = cols, pch = pchs, cex = cexs,
      xlab = "Measurement 1", ylab = "Measurement 2", zlab = "M 3",
      screen = cam)

cloud(pc3 ~ pc1 * pc2, data = combined_rot, groups = grp, col = cols, pch = pchs, cex = cexs,
      xlab = "PC 1", ylab = "PC 2", zlab = "PC 3",
      screen = cam)

ggplot(rot3, aes(pc1, pc2, color = group)) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_manual(values = pal) +
  labs(x = "PC 1", y = "PC 2", color = NULL) +
  theme_minimal()
```

## MDS: Building a Map from Distances Alone

MDS starts from a table of how different every pair of samples is from every other pair, and then searches for an arrangement of dots on a page where the distances between dots match those real differences as closely as possible. It never needs to see the original measurements — only how different each pair of samples is.

Here's that same process end to end, with a small, realistic example. The 10 genes below are commonly checked for promoter methylation in cancer — when a tumor suppressor like *TP53* or *PTEN* gets silenced this way, the cell loses one of its safeguards against uncontrolled growth. We have percent methylation at each gene, in 3 samples.

```{r}
#| label: tbl-epigenetic-data
#| tbl-cap: "Percent DNA methylation at 10 genes, in 3 samples. This is the kind of raw table a report usually starts from."

library(gt)
suppressPackageStartupMessages(library(dplyr))

set.seed(7)
genes <- c("BRCA1", "TP53", "MLH1", "MGMT", "CDKN2A", "RASSF1", "APC", "VHL", "PTEN", "RB1")
sample_a <- round(runif(10, 20, 40), 1)
sample_b <- round(sample_a + rnorm(10, 0, 3), 1)
sample_c <- round(runif(10, 55, 85), 1)

meth <- data.frame(
  Gene = genes,
  `Sample A` = sample_a,
  `Sample B` = sample_b,
  `Sample C` = sample_c,
  check.names = FALSE
)

meth %>%
  gt() %>%
  tab_header(title = "Percent DNA Methylation at 10 Genes") %>%
  fmt_number(columns = -Gene, decimals = 1) %>%
  opt_row_striping()
```

The first step MDS takes is to boil that whole table down to a single number per pair of samples — how different is Sample A from Sample B, overall, across all 10 genes at once?

```{r}
#| label: tbl-epigenetic-distances
#| tbl-cap: "How different each pair of samples is, considering all 10 genes together. Lower means more alike."

m <- as.matrix(meth[, -1])
rownames(m) <- meth$Gene
d <- dist(t(m))
dm <- as.data.frame(as.matrix(d))
dm <- cbind(Sample = rownames(dm), dm)

dm %>%
  gt() %>%
  tab_header(title = "Overall Difference Between Samples") %>%
  fmt_number(columns = -Sample, decimals = 1) %>%
  opt_row_striping()
```

That table alone is all MDS needs. It has no idea these numbers came from 10 genes, or even that "genes" were involved at all — it only sees three numbers describing how far apart each pair is. From just that, it draws the map:

```{r}
#| label: fig-epigenetic-mds
#| fig-cap: "MDS map built entirely from the distance table above. Samples A and B, the two closest pairs in the table, land near each other; Sample C, far from both, lands on its own."

library(ggplot2)

mds_toy <- as.data.frame(cmdscale(d, k = 2))
names(mds_toy) <- c("dim1", "dim2")
mds_toy$sample <- colnames(m)

ggplot(mds_toy, aes(dim1, dim2, label = sample)) +
  geom_point(size = 3.5, color = "#4e79a7") +
  geom_text(vjust = -1.3, size = 4.5) +
  scale_x_continuous(expand = expansion(mult = 0.35)) +
  scale_y_continuous(expand = expansion(mult = 0.35)) +
  labs(x = "MDS Dimension 1", y = "MDS Dimension 2") +
  theme_minimal()
```

Samples A and B — the pair with the smallest number in the distance table — land close together on the map. Sample C, far from both in the table, lands off on its own. Nothing here required looking at which specific genes drove that difference; the map is built purely from the one column of numbers in between the raw data and the picture.

## So What's the Actual Difference?

- **PCA** starts from the original measurements and looks for the directions that show the most variation between samples.
- **MDS** starts from a table of pairwise differences and tries to reproduce those differences as distances on a flat page.

In many everyday cases, these two approaches end up producing very similar-looking plots — "the view that shows the most variation" and "the layout that best preserves how different everything is" often point to the same picture. 

Here's that similarity in practice. The toy data below isn't two-dimensional to begin with — each of the 30 samples has **8 separate measurements**, impossible to plot directly. The same 8-measurement data was handed to both PCA and MDS independently, with no coordination between them:

```{r}
#| label: fig-pca-mds-compare
#| fig-cap: "The same 30 samples, each with 8 underlying measurements, flattened to 2D by PCA (left) and MDS (right). Both methods were given the exact same data and arrived at essentially the same picture: three clusters and one outlier."
#| fig-subcap:
#|   - "PCA"
#|   - "MDS"
#| layout-ncol: 2

library(ggplot2)

set.seed(11)
n_per_group <- 10
n_vars <- 8

# each group's signal lives in its own non-overlapping set of the 8 measurements,
# so the three groups are equally, cleanly separated
center_A <- c(6, 6, 6, 0, 0, 0, 0, 0)
center_B <- c(0, 0, 0, 6, 6, 6, 0, 0)
center_C <- c(0, 0, 0, 0, 0, 0, 6, 6)

make_group <- function(center) {
  sweep(matrix(rnorm(n_per_group * n_vars), nrow = n_per_group), 2, center, "+")
}

raw <- rbind(make_group(center_A), make_group(center_B), make_group(center_C))
group <- rep(c("Group A", "Group B", "Group C"), each = n_per_group)

# one sample, off across every measurement -- a true outlier, not just a 2D one
raw[1, ] <- raw[1, ] + 10

pal <- c("Group A" = "#e15759", "Group B" = "#59a14f", "Group C" = "#4e79a7")

pca <- prcomp(raw, scale. = TRUE)
pca_df <- data.frame(dim1 = pca$x[, 1], dim2 = pca$x[, 2], group = group)

mds <- cmdscale(dist(raw), k = 2)
mds_df <- data.frame(dim1 = mds[, 1], dim2 = mds[, 2], group = group)

ggplot(pca_df, aes(dim1, dim2, color = group)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = pal) +
  labs(x = "PC 1", y = "PC 2", color = NULL) +
  theme_minimal()

ggplot(mds_df, aes(dim1, dim2, color = group)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = pal) +
  labs(x = "MDS Dimension 1", y = "MDS Dimension 2", color = NULL) +
  theme_minimal()
```

## What to Look Fot

- **Clusters** — groups of dots sitting close together are samples that look broadly alike across everything measured.
- **A dot sitting off on its own** — an outlier, worth a second look. It might be a data quality issue, or it might be genuinely unusual.
- **Two groups separating cleanly along one direction** — a real, dominant difference between them, worth asking what drove it.
- **Everything blending into one shapeless cloud** — no strong pattern in what was measured, at least not one big enough to show up this way.
